CDI Schubert

### 2017-09-29 datasets/microbiome/cdi_schubert.Rmd
Add Goodrich and diagnostic plots

```{r}
library(dplyr)
library(ggplot2)

setwd("/Users/claire/github/benchmark-fdr/datasets/microbiome")
source("../../simulation_studies/R/simulation-helpers.R")
source("../R/evaluation-helpers.R")
resdir <- "/Users/claire/github/benchmark-fdr/datasets/microbiome/"
```

@@

```{r, results='hide'}
# load in dbOTU table.
otu <- read.table('cdi_schubert_results/RDP/cdi_schubert.otu_table.dbOTU.rdp_assigned')
# and metadata
meta <- read.csv('cdi_schubert_results/cdi_schubert.metadata.txt', sep='\t')
head(otu)
```

```{r, results='hide'}
## Clean up the data

# Keep only samples with DiseaseState metadata
labels <- c("H", "nonCDI", "CDI")

print(dim(meta))
meta <- meta %>% filter(DiseaseState %in% labels)
print(dim(meta))

# Metadata and 16S data
keep_samples <- intersect(colnames(otu), meta$sample_id)
otu <- otu[, keep_samples]
meta <- meta %>% filter(sample_id %in% keep_samples)
print(dim(otu))
print(dim(meta))

# Remove OTUs and samples without enough reads
n_reads_samples <- 1000
n_reads_otus <- 10

@@

print(dim(otu))
otu <- remove_shallow_smpls(otu, n_reads_samples)
otu <- remove_shallow_otus(otu, n_reads_otus)
print(dim(otu))

# Update metadata with new samples
meta <- meta %>% filter(sample_id %in% colnames(otu))

```{r, results='hide'}
## Calculate some covariates

@@

# Mean relative abundance (only considering people who have the OTU present)
abun_otu <- t(t(otu) / rowSums(t(otu)))
mean_abun <- rowSums(abun_otu) / rowSums(abun_otu > 0)

# Make new dataframe to store covariates and eventual p-values
schubert <- data.frame(otu = rownames(otu), mean_abun = mean_abun,
                       ubiquity = ubi, mean_reads = reads)
head(schubert)
```

```{r}
## Calculate p-values between case (CDI or nonCDI) and controls (H)
## Use the relative abundances!

# just do a for loop and ask how to vectorize l8er
case_idx <- meta$DiseaseState %in% c("CDI", "nonCDI")
ctrl_idx <- meta$DiseaseState %in% c("H")
pvals <- c()
ses <- c()
effs <- c()
for (o in rownames(abun_otu)) {
  p <- wilcox.test(abun_otu[o, ctrl_idx], 
                   abun_otu[o, case_idx])$p.value
  pvals <- c(pvals, p)

@@

  # effect (difference of means)
  effs <- c(effs, mean(abun_otu[o, ctrl_idx]) - mean(abun_otu[o, case_idx]))
}
```

```{r}
schubert$pval <- pvals
schubert$SE <- ses
schubert$effect_size <- effs

# Set up ubiquity as the first covariate to test.
# Ranges from 0.001 to 0.4
schubert <- schubert %>% 
              mutate(zscore = effect_size/SE) %>%
              mutate(ind_covariate = log10(ubiquity) )
```

```{r, results='hide', eval = FALSE}
head(schubert)
```

```{r}
# Start with ubiquity as the covariate
# TODO: check independence under the null

# Copied from Keegan's code :)
cutoffs <- c(0.01, 0.025, 0.05, 0.10, 0.20)
resfile <- paste0(resdir, "cdi_schubert_results_",
                  nrow(schubert), ".RData")
t1 <- proc.time()
schu_results <- run_benchmarks(dat=schubert, alphas=cutoffs)
save(schu_results, file=resfile)
message(paste0("Took ", round((proc.time()-t1)[3],1),
               " seconds for ", nrow(schubert), " OTUs"))
```

```{r}
# plot % rejects vs FDR cutoff
schu_results <- schu_results %>% mutate(perc_reject = n_rejects / length(rownames(abun_otu)))
p <- ggplot(schu_results, aes(alpha, perc_reject, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Fraction of hypotheses rejected") +
  ggtitle("CDI Schubert with log10(Ubiquity) Covariate") +
  theme_classic()
print(p)

@@

>>>>

```{r}
library(dplyr)
library(ggplot2)
library(cowplot)

wd <- "/Users/claire/github/benchmark-fdr/datasets/microbiome/"
setwd(wd)
source("../../simulation_studies/R/simulation-helpers.R")
source("../R/evaluation-helpers.R")
resdir <- "/Users/claire/github/benchmark-fdr/datasets/microbiome/"
```

@@

```{r, data_functions, echo = FALSE, results='hide'}
# Define functions to read and clean up data

@@

read_and_clean_data <- function(otu_path, meta_path, labels, col_label,
                                sample_reads = 1000, otu_reads = 10) {
  # Reads the OTU table and metadata from otu_path and meta_path,
  # keeps samples which have both metadata and 16S data and whose
  # DiseaseState is in the list labels. Removes samples with fewer
  # than sample_reads reads, and and OTUs with fewer than otu_reads
  # reads. Metadata should have sample IDs in first column and a column
  # labeled DiseaseState

  # load OTU table and metadata
  otu <- read.table(otu_path)
  meta <- read.csv(meta_path, sep='\t')

  # Keep only samples with the right DiseaseState metadata
  meta <- meta %>% filter(DiseaseState %in% labels)

  # Keep only samples with both metadata and 16S data
  keep_samples <- intersect(colnames(otu), meta[, 1])
  otu <- otu[, keep_samples]
  meta <- meta %>% filter(get(col_label) %in% keep_samples)
  # Other ways to do this:
  #meta[meta[, col_label] %in% keep_samples, ]
  #meta[col_label]
  #`[[`(meta, col_label)

  # Remove OTUs and samples without enough reads
  otu <- remove_shallow_smpls(otu, sample_reads)
  otu <- remove_shallow_otus(otu, otu_reads)

  # Update metadata with new samples
  meta <- meta %>% filter(get(col_label) %in% colnames(otu))

  return(list(otu, meta))
}
```

```{r, covariate_calcs, echo = FALSE, results='hide'}
calculate_covariates <- function(otu, abun_otu){
  # Calculates covariates for the OTU table.
  # OTU table should have OTUs in rows and samples in columns.
  # This function calculates (per OTU): mean read depth, 
  # ubiquity, mean relative abundance in non-zero samples.
  # It returns a dataframe with each of these as columns, in addition
  # to "otu".
  # zeroabun is the value in abun_otu corresponding to zero abundance.

@@

  # Mean relative abundance (only considering people who have the OTU present)
  mean_abun <- rowSums(abun_otu) / rowSums(abun_otu > 0)

  covdf <- data.frame(otu = rownames(otu), mean_abun = mean_abun,
                      ubiquity = ubi, mean_reads = reads)
  return(covdf)
}
```

```{r, calculate_pvals, echo = FALSE}
calculate_pvals <- function(abun_otu, ctrl_idx, case_idx){
  # Does wilcoxon rank sum test for all OTUs between cases and
  # controls. abun_otu is a dataframe with OTUs in rows and 
  # samples in columns. ctrl_idx and case_idx are boolean
  # vectors indicating the columns correspnding to control
  # and case samples, respectively.
  # This function also calculate the effect size 
  # (difference of mean abundances) and standard error

  pvals <- c()
  ses <- c()
  effs <- c()

  casedf <- abun_otu[, case_idx]
  ctrldf <- abun_otu[, ctrl_idx]
  for (o in rownames(abun_otu)) {
    p <- wilcox.test(casedf[o, ], 
                     ctrldf[o, ])$p.value
    pvals <- c(pvals, p)

@@

    # effect (difference of means)
    effs <- c(effs, mean(abun_otu[o, ctrl_idx]) - mean(abun_otu[o, case_idx]))

  }
  res <- data.frame(otu = rownames(abun_otu), pval = pvals, SE = ses, effect_size = effs)
  return(res)
}

@@

```{r, def_strat_hist}
strat_hist <- function(dat, pval, covariate, binwidth=0.025, maxy=3){
  gg_all <- ggplot(data=schubertdat, aes(x=get(pcol))) + 
    geom_histogram(binwidth = binwidth, boundary = 0, 
                   colour="grey", fill="lightgrey") +
    aes(y=..density..)+
    theme_classic() +
    theme(axis.title = element_text(face="bold"),
          plot.title = element_text(face="bold")) +
    scale_x_continuous(expand = c(0.02, 0)) + 
    scale_y_continuous(expand = c(0.02, 0), limits=c(0,maxy)) + 
    xlab("p-value") +
    ylab("Density") +
    ggtitle("All OTUs")

  dat.strat <- schubertdat %>% 
      filter(-rank(get(covcol), ties="first") > quantile(-rank(get(covcol), ties="first"), 2/3))
  gg_top <- ggplot(data=dat.strat, aes(x=get(pcol))) + 
    geom_histogram(binwidth = binwidth, boundary = 0, 
                   colour="grey", fill="lightgrey") +
    aes(y=..density..) +
    theme_classic() +
    theme(axis.title = element_text(face="bold"),
          plot.title = element_text(face="bold")) +
    scale_x_continuous(expand = c(0.02, 0)) + 
    scale_y_continuous(expand = c(0.02, 0), limits=c(0,maxy)) + 
    xlab("p-value") +
    ylab("Density") +
    ggtitle(paste0("Top third by ", covcol))

  dat.strat <- dat %>% 
    filter(-rank(get(covcol), ties="first") <= quantile(-rank(get(covcol), ties="first"), 2/3) &
           -rank(get(covcol), ties="first") > quantile(-rank(get(covcol), ties="first"), 1/3) )
  gg_mid <- ggplot(data=dat.strat, aes(x=get(pcol))) + 
    geom_histogram(binwidth = binwidth, boundary = 0, 
                   colour="grey", fill="lightgrey") +
    aes(y=..density..)+
    theme_classic() +
    theme(axis.title = element_text(face="bold"),
          plot.title = element_text(face="bold")) +
    scale_x_continuous(expand = c(0.02, 0)) + 
    scale_y_continuous(expand = c(0.02, 0), limits=c(0,maxy)) + 
    xlab("p-value") +
    ylab("Density") +
    ggtitle(paste0("Middle third by ", covcol))

  dat.strat <- dat %>% 
    filter(-rank(get(covcol), ties="first") <= quantile(-rank(get(covcol), ties="first"), 1/3))
  gg_bot <- ggplot(data=dat.strat, aes(x=get(pcol))) + 
    geom_histogram(binwidth = binwidth, boundary = 0, 
                   colour="grey", fill="lightgrey") +
    aes(y=..density..) +
    theme_classic() +
    theme(axis.title = element_text(face="bold"),
          plot.title = element_text(face="bold")) +
    scale_x_continuous(expand = c(0.02, 0)) + 
    scale_y_continuous(expand = c(0.02, 0), limits=c(0,maxy)) + 
    xlab("p-value") +
    ylab("Density") +
    ggtitle(paste0("Lower third by ", covcol))

  gg_stratified <- plot_grid(gg_all, gg_top, gg_mid, gg_bot,
                             nrow=1,
                             labels=c("(a)", "(b)", "(c)","(d)"),
                             hjust = 0.1)
  return(gg_stratified)
}
```

```{r}
# Read and clean cdi_schubert
otu_path = paste0(wd, "cdi_schubert_results/RDP/cdi_schubert.otu_table.100.denovo.rdp_assigned")
meta_path = paste0(wd, "cdi_schubert_results/cdi_schubert.metadata.txt")
labels = c("H", "nonCDI", "CDI")

dat <- read_and_clean_data(otu_path, meta_path, labels, 'sample_id')
otu <- dat[[1]]
meta <- dat[[2]]

abun_otu <- t(t(otu) / rowSums(t(otu)))
minabun <- min(abun_otu[abun_otu > 0])
zeroabun = 1e-6
abun_otu <- abun_otu + zeroabun
```

```{r, schubert-pvals}
case_idx <- meta$DiseaseState %in% c("CDI", "nonCDI")
ctrl_idx <- meta$DiseaseState %in% c("H")

# Calculate some basic covariates for each OTU
schubertdat <- calculate_covariates(otu, abun_otu)
# Get the pvalues, effects, and SEs
schu_pvals <- calculate_pvals(abun_otu, ctrl_idx, case_idx)
schubertdat <- merge(schubertdat, schu_pvals, by="otu")
schubertdat <- schubertdat %>% 
                 mutate(logpval = log10(pval)) %>%
                 mutate(neglogpval = -log10(pval)) %>%
                 mutate(logubiquity = log10(ubiquity)) %>%
                 mutate(logmean_reads = log10(mean_reads))
head(schubertdat)
```

```{r}
pcol <- "pval"
covcol <- "logubiquity"
binwidth <- 0.025
maxy <- 3
strat_hist(schubertdat, pcol, covcol)
```
```{r}
g <- ggplot(schubertdat, aes(x = get(covcol), y = pval)) + geom_point()
g 
```

```{r}
covcol <- "logmean_reads"
strat_hist(schubertdat, pcol, covcol)
```

```{r}
g <- ggplot(schubertdat, aes(x = get(covcol), y = pval)) + geom_point()
g 
```

```{r, run_cdi_schubert}
# Set up ubiquity as the first covariate to test.
# Ranges from 0.001 to 0.4
schubertdat <- schubertdat %>% 
              mutate(zscore = effect_size/SE) %>%
              mutate(ind_covariate = log10(ubiquity) )

# Run all the tests and plot
cutoffs <- c(0.01, 0.025, 0.05, 0.10, 0.20)
resfile <- paste0(resdir, "cdi_schubert_results_",
                  nrow(schubertdat), ".RData")
t1 <- proc.time()
schu_results <- run_benchmarks(dat=schubertdat, alphas=cutoffs)
save(schu_results, file=resfile)
message(paste0("Took ", round((proc.time()-t1)[3],1),
               " seconds for ", nrow(schubertdat), " OTUs"))

# plot % rejects vs FDR cutoff
schu_results <- schu_results %>% mutate(perc_reject = n_rejects / length(rownames(abun_otu)))
p <- ggplot(schu_results, aes(alpha, perc_reject, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Fraction of hypotheses rejected") +
  ggtitle("CDI Schubert with log10(Ubiquity) Covariate") +
  theme_classic()
print(p)
```

```{bash, goodrich-download, eval=FALSE}
curl -O "https://zenodo.org/record/840333/files/ob_goodrich_results.tar.gz"
tar -xzvf ob_goodrich_results.tar.gz
```

```{r, read-ob-goodrich}
# Read and clean ob_goodrich
otu_path = paste0(wd, "ob_goodrich_results/RDP/ob_goodrich.otu_table.100.denovo.rdp_assigned")
meta_path = paste0(wd, "ob_goodrich_results/ob_goodrich.metadata.txt")
labels = c("H", "OB")

dat <- read_and_clean_data(otu_path, meta_path, labels, "X")
otu <- dat[[1]]
meta <- dat[[2]]

abun_otu <- t(t(otu) / rowSums(t(otu)))
minabun <- min(abun_otu[abun_otu > 0])
zeroabun = 1e-6
abun_otu <- abun_otu + zeroabun
```

```{r, goodrich-pvals}
case_idx <- meta$DiseaseState %in% c("OB")
ctrl_idx <- meta$DiseaseState %in% c("H")

# Calculate some basic covariates for each OTU
goodrichdat <- calculate_covariates(otu, abun_otu)
# Get the pvalues, effects, and SEs
goodrich_pvals <- calculate_pvals(abun_otu, ctrl_idx, case_idx)
goodrichdat <- merge(goodrichdat, goodrich_pvals, by="otu")
```

```{r}
goodrichdat <- goodrichdat %>% 
                 mutate(logpval = log10(pval)) %>%
                 mutate(neglogpval = -log10(pval)) %>%
                 mutate(logubiquity = log10(ubiquity)) %>%
                 mutate(logmean_reads = log10(mean_reads))
head(goodrichdat)
```

```{r}
covcol <- "logubiquity"
strat_hist(goodrichdat, pcol, covcol)
```

```{r, explore-goodrich-covariates}
var <- "logubiquity"
g <- ggplot(goodrichdat, aes(x = get(var), y = pval)) + geom_point()
g
```

```{r}
covcol <- "logmean_reads"
strat_hist(goodrichdat, pcol, covcol)
```

```{r}
g <- ggplot(goodrichdat, aes(x = get(covcol), y = pval)) + geom_point()
g
```

```{r, test-fdrs-ubiquity}
# Set up ubiquity as the first covariate to test.
# Ranges from 0.001 to 0.4
goodrichdat <- goodrichdat %>% 
              mutate(zscore = effect_size/SE) %>%
              mutate(ind_covariate = log10(ubiquity) )

# Run all the tests and plot
cutoffs <- c(0.01, 0.025, 0.05, 0.10, 0.20)
resfile <- paste0(resdir, "ob_goodrich_results_",
                  nrow(goodrichdat), ".RData")
t1 <- proc.time()
goodrich_results <- run_benchmarks(dat=goodrichdat, alphas=cutoffs)
save(goodrich_results, file=resfile)
message(paste0("Took ", round((proc.time()-t1)[3],1),
               " seconds for ", nrow(goodrichdat), " OTUs"))

# plot % rejects vs FDR cutoff
goodrich_results <- goodrich_results %>% 
  mutate(perc_reject = n_rejects / length(rownames(abun_otu)))

p <- ggplot(goodrich_results, aes(alpha, perc_reject, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Fraction of hypotheses rejected") +
  ggtitle("OB Goodrich with log10(Ubiquity) Covariate") +
  theme_classic()
print(p)
```
